Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-07-08 summary.tex
updated document

We will use the notation \(A \sim \PABM(\{\lambda^{(kl)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\lambda^{(kl)}\) consisting of \(K\) underlying communities.

@@

@@

\begin{remark}
Like the RDPG, the latent positions of a GRDPG are not unique
\cite{rubindelanchy2017statistical}.
More specifically, if $P_{ij} = x_i^\top I_{p, q} x_j$, then we also have for any
$Q \in \mathbb{O}(p, q)$,
$(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j =
x_i^\top I_{p, q} x_j = P_{ij}$.
Unlike in the RDPG case, transforming the latent positions via multiplication
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or
distances.
\end{remark}

\hypertarget{connecting-the-pabm-to-the-grdpg}{%
\subsection{Connecting the PABM to the
GRDPG}\label{connecting-the-pabm-to-the-grdpg}}
\begin{proposition}[Connecting the PABM to the GRDPG for $K = 2$]
\label{theorem1}
Let
$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix} \quad \text{and} \quad
U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
0 & 1 & 0 & 0 \end{bmatrix},$$
where the $\{\lambda^{(kl)}\}$ are as defined in Definition 1.
Then $A \sim \mathrm{GRDPG}_{3, 1}(X U)$ and $B \sim \mathrm{PABM}(\{(\lambda^{(kl)}\})$ are
identically distributed.
\end{proposition}

\begin{proof}
  We can show that
\[X U I_{3, 1} U^\top X^\top =
\begin{bmatrix}
  \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix} \]
by straightforward matrix multiplication. It is nevertheless
instructive to look at a few intermediate steps. More specifically,
the product $U I_{3, 1} U^\top$ yields a permutation matrix $\Pi$
with fixed points at positions $1$ and $4$ and a cycle of order 2 swapping
positions $2$ and $3$, i.e., 
$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 1
\end{bmatrix}.$$
Furthermore, since $U$ is orthonormal and $I_{3, 1}$ is diagonal, $U I_{3, 1} U^\top$
is also an eigendecomposition of $\Pi$ where the fixed
points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
while the cycles of order two are mapped to the eigenvectors  
$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
e_3)$; here $e_i$ denote the $i$th basis vector in $\mathbb{R}^{4}$. %  and $ eigenvalues $\pm 1$ with corresponding
% eigenvectors $(e_i \pm e_j) / \sqrt{2}$ where $i = 2, j = 3$, pair that is
% swapped.
\end{proof}
% The form for the matrix $\Pi$ used in the proof of
% Proposition~\ref{theorem1} can 
\begin{theorem}[Generalization to $K > 2$]
\label{theorem2}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\})$ be an instance of a
PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. Then there exists a block diagonal matrix
$X \in \mathbb{R}^{n \times K^2}$ defined by the $\{\lambda^{(kl)}\}$ and a $K^2 \times
K^2$ {\em fixed} orthonormal matrix $U$ such 
that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) / 2}(XU)$.
\end{theorem}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Community membership identification (up to permutation).
\item
  Parameter estimation (estimating \(\lambda^{(kl)}\)'s).
\end{enumerate}

@@

\citet*{noroozi2019estimation} used Sparse Subspace Clustering (SSC) for
community detection in the PABM. SSC is performed by solving an
optimization problem for each observed point. Given
\(X \in \mathbb{R}^{n \times d}\) with vectors
\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
\(c_i = \arg\min_{c} ||c||_1\) subject to \(x_i = X c\) and
\(c^{(i)} = 0\) is solved for each \(i = 1, ..., n\). The solutions are
collected collected into matrix
\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
construct an affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
perfectly on one of \(K\) subspaces, \(B\) describes an undirected graph
consisting of \(K\) disjoint subgraphs, i.e., \(B_{ij} = 0\) if
\(x_i, x_j\) are in different subspaces. If \(X\) instead represents
points near \(K\) subspaces with some noise, a final graph partitioning
step may be performed (e.g., edge thresholding or spectral clustering).

@@

For any edge probability matrix \(P\) for the PABM such that the rows
and columns are organized by community, the \(kl\)\textsuperscript{th}
block is an outer product of two vectors, i.e.,
\(P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top\). Therefore, given
\(P^{(kl)}\), \(\lambda^{(kl)}\) and \(\lambda^{(lk)}\) are solvable
up to multiplicative constant using singular value
decomposition. More specifically, 
let \(P^{(kl)} = (\sigma^{(kl)})^2 u^{(kl)} (v^{(kl)})^\top\)
be the singular value decomposition of \(P^{(kl)}\).
\(u^{(kl)} \in \mathbb{R}^{n_k}\) and 
\(v^{(kl)} \in \mathbb{R}^{n_l}\) are vectors
and \(\sigma^{(kl)}\) is a scalar. Then \(\lambda^{(kl)} = s_1 u^{(kl)}\)
and \(\lambda^{(lk)} = s_2 v^{(kl)}\) 
for unidentifiable $s_1 s_2 = (\sigma^{(kl)})^2$.
Given the adjacency matrix \(A\)
instead of edge probability matrix \(P\), we can simply use plug-in
estimators by taking the SVD of each $A^{(kl)}$. 
Since each $\lambda^{(kl)}$ is not strictly identifiable,
we instead estimate each $\tilde{\lambda}^{(kl)} = \sigma^{(kl)} u^{(kl)}$.

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation.}
  \KwData{Adjacency matrix $A$, community assignments $1, ..., K$}
  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(kl)}\}_K$.}
  Arrange the rows and columns of $A$ by community such that each 
  $A^{(kl)}$ block consists of estimated edge probabilities between 
  communities $k$ and $l$.\;
  \For {$k, l = 1, ..., K$, $k \leq l$} {
    Compute $A^{(kl)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ 
    block.\;
    Assign $u^{(kl)}$ and $v^{(kl)}$ as the first columns of $U$ and $V$. 
    Assign $(\sigma^{(kl)})^2 \leftarrow \Sigma_{11}$.\;
    Assign $\hat{\lambda}^{(kl)} \leftarrow \pm \sigma^{(kl)} u^{(kl)}$ and 
    $\hat{\lambda}^{(lk)} \leftarrow \pm \sigma^{(kl)} v^{(kl)}$.
  }
\end{algorithm}

\begin{theorem}
\label{theorem6}
Let each $\tilde{\lambda}^{(kl)}$ be the popularity vector derived from its corresponding $P^{(kl)}$, and let $\hat{\lambda}^{(kl)}$ be its estimate from $A^{(kl)}$ (Alg. 3). Then if $n \rho_n = \omega \big( (\log n)^{4 c} \big)$,

\begin{equation} \label{eq:thm6}
\max_{k, l \in \{1, ..., K\}} 
\|\hat{\lambda}^{(kl)} - \tilde{\lambda}^{(kl)}\|_{\infty} = 
O\bigg(\frac{(\log n_k)^c}{\sqrt{n_k}} \bigg)
\end{equation}

@@

For each simulation, community labels are drawn from a multinomial
distribution, the popularity vectors \(\{\lambda^{(kl)}\}_K\) are drawn
from two types of joint distributions depending on whether \(k = l\),
the edge probability matrix \(P\) is constructed using the popularity
vectors, and finally an unweighted and undirected adjacency matrix \(A\)
is drawn from \(P\). OSC is then used for community detection, and this
method is compared against SSC \cite{noroozi2019estimation,soltanolkotabi2014}
and MM \cite{igraph, 307cbeb9b1be48299388437423d94bf1}. 
True community labels are used
with Algorithm 3 to estimate the popularity vectors
\(\{\lambda^{(kl)}\}_K\), and this method is then compared against an
MLE-based estimator described by \citet{noroozi2019estimation} and
\citet{307cbeb9b1be48299388437423d94bf1}.

@@

We also define two types of parameter estimation error. 
First, we estimate the popularity vectors directly and compute the RMSE. 
In this case, the "true" popularity vectors are derived from 
taking the SVD of each edge probability block $P^{(kl)}$ 
to avoid the unidentifiable multiplicative constants.

\[\text{RMSE}(\{\hat{\lambda}^{(kl)}\}_K, \{\lambda^{(kl)}\}_K) = 
\sqrt{\frac{1}{n} \sum_{k < l} \min_{s = \pm 1} 
\|s \hat{\lambda}^{(kl)} - \lambda^{(kl)} \|_2^2}\]

We can also avoid the unidentifiable multiplicative constant more directly 
by reconstructing each 
$\hat{P}^{(kl)} = \hat{\lambda}^{(kl)} (\hat{\lambda}^{(lk)})^\top$, 
which we use to define another parameter estimation error.

\[\text{RMSE}(\hat{P}, P) =
\sum_{k,l} \sqrt{\frac{1}{n_k n_l} \|P^{(kl)} - \hat{P}^{(kl)}\|_F^2}\]

@@

In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
multinomial distribution with mixture parameters
\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(kl)}\}_K\) according
to the drawn community labels, \(P\) was constructed using the drawn
\(\{\lambda^{(kl)}\}_K\), and \(A\) was drawn from \(P\) by
\(A_{ij} \indep \Bernoulli(P_{ij})\). Each simulation has
a unique edge probability matrix \(P\).

@@

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = 1 / K\) for \(k = 1, ..., K\), (i.e.,
  each community label has an equal probability of being drawn)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(kl)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
  \(k \neq l\)
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = \frac{k^{-1}}{\sum_{l=1}^K l^{-1}}\)
  for \(k = 1, ..., K\)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(kl)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
  \(k \neq l\)
\end{itemize}

@@

\hypertarget{proofs}{%
\section{Proofs of Theorem~\ref{theorem4} and Theorem~\ref{theorem5}}}
Let \(V_n\) and \(\hat{V}_n\)
be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
$K^2$ largest eigenvalues (in modulus), respectively. 
We first state an important technical lemma for bounding the maximum
$\ell_2$ norm difference between the rows of $\hat{V}_n$ and
$V_n$. See \cite{cape_biometrika} and 
\cite[Lemma~5]{rubindelanchy2017statistical} for a proof. 
\begin{lemma}
  \label{lem:technical}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\}_{K})$ be a $K$-blocks
PABM graph on $n$ vertices and let \(V_n\) and \(\hat{V}_n\)
be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
$K^2$ largest eigenvalues in modulus, respectively.
Let \((v_n^{(i)})^\top\) and \((\hat{v}_n^{(i)})^\top\) denote the $i$th 
row of \(V_n\) and \(\hat{V}_n\), respectively. 
Then there exists a constant $c > 1$ and an orthogonal matrix $W$ such
that, with high probability,
$$\max_{i} \|W \hat{v}_n^{(i)} - v_n^{(i)}\|  = O_P \Big(\frac{\log^{c}n}{n \sqrt{\rho_n}} \Big).$$
In particular we can take $c = 1 + \epsilon$ for any $\epsilon > 0$. 
\end{lemma}

@@

which is a symmetric \((n_k + n_l) \times (n_k + n_l)\) matrix. The
eigendecomposition of \(M^{(kl)}\) is then

\[M^{(kl)} = 
\begin{bmatrix} u^{(k \ell)} & -u^{(k \ell)} \\ v^{(k \ell)} & v^{(k \ell)} \end{bmatrix} \times 
\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
\begin{bmatrix} u^{(k \ell)} & -u^{(k \ell)} \\ v^{(k \ell)} & v^{(k \ell)} \end{bmatrix}^\top\]
Thus treating \(M^{(k \ell)}\) as the edge probability matrix of a GRDPG, we
have latent positions in \(\mathbb{R}^2\) given by the $(n_k + n_{\ell}) \times 2$ matrix
\[\Lambda^{(k \ell)} = \begin{bmatrix} 
  \sigma_{k \ell} u^{(k \ell)} & \sigma_{k \ell} u^{(k \ell)} \\ 
  \sigma_{k \ell} v^{(k \ell)} & -\sigma_{k \ell} v^{(k \ell)} 
\end{bmatrix} = 
\begin{bmatrix} 
  \lambda^{(k \ell)} & \lambda^{(k \ell)} \\ 
  \lambda^{( \ell k)} & -\lambda^{( \ell k)} 
\end{bmatrix}.\]
Now consider
\[\hat{M}^{(k \ell)} = \begin{bmatrix} 0 & A^{(k \ell)} \\ A^{(\ell k)} & 0 \end{bmatrix}\]
% Then \(\hat{M}^{(kl)} = M^{(kl)} + E'\) where

>>>>

We will use the notation \(A \sim \PABM(\{\lambda^{(k \ell)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\lambda^{(k \ell)}\) consisting of \(K\) underlying communities.

@@

\begin{remark}
We can use Adjacency Spectral Embedding (ASE) 
\cite{rubindelanchy2017statistical} to recover the latent vectors of a GRDPG. 
This procedure consists of taking the spectral decomposition of $A$ 
(or $P$ if available) and choosing the $p$ most positive and $q$ most negative
eigenvalues and their corresponding eigenvectors 
to construct the embedding $\hat{Z} = V |D|^{1/2}$. 
\end{remark}

@@

\begin{remark}
The latent vectors that produce $X I_{p,q} X^\top = P$ are not unique
\cite{rubindelanchy2017statistical}.
More specifically, if $P_{ij} = x_i^\top I_{p, q} x_j$, 
then we also have for any$Q \in \mathbb{O}(p, q)$,
$(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j =
x_i^\top I_{p, q} x_j = P_{ij}$.
Unlike in the RDPG case, transforming the latent positions via multiplication
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or
distances.
\end{remark}

\hypertarget{connecting-the-pabm-to-the-grdpg}{%
\subsection{Connecting the PABM to the
GRDPG}\label{connecting-the-pabm-to-the-grdpg}}
\begin{proposition}[Connecting the PABM to the GRDPG for $K = 2$]
\label{theorem1}
Let
$$X = \begin{bmatrix}
\lambda^{(11)} & \lambda^{(12)} & 0 & 0 \\
0 & 0 & \lambda^{(21)} & \lambda^{(22)}
\end{bmatrix} \quad \text{and} \quad
U = \begin{bmatrix} 1 & 0 & 0 & 0 \\
0 & 0 & 1 / \sqrt{2} & 1 / \sqrt{2} \\
0 & 0 & 1 / \sqrt{2} & - 1 / \sqrt{2} \\
0 & 1 & 0 & 0 \end{bmatrix},$$
where the $\{\lambda^{(k \ell)}\}$ are as defined in Definition 1.
Then $A \sim \mathrm{GRDPG}_{3, 1}(X U)$ and 
$B \sim \mathrm{PABM}(\{(\lambda^{(k \ell)}\}_K)$ are
identically distributed.
\end{proposition}

\begin{proof}
  We can show that
\[X U I_{3, 1} U^\top X^\top =
\begin{bmatrix}
  \lambda^{(11)} (\lambda^{(11)})^\top & \lambda^{(12)} (\lambda^{(21)})^\top \\
  \lambda^{(21)} (\lambda^{(12)})^\top & \lambda^{(22)} (\lambda^{(22)})^\top
\end{bmatrix} \]
by straightforward matrix multiplication. It is nevertheless
instructive to look at a few intermediate steps. More specifically,
the product $U I_{3, 1} U^\top$ yields a permutation matrix $\Pi$
with fixed points at positions $1$ and $4$ and a cycle of order 2 swapping
positions $2$ and $3$, i.e., 
$$\Pi = U I_{3, 1} U^\top = \begin{bmatrix} 1 & 0 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 0 & 1
\end{bmatrix}.$$
Furthermore, since $U$ is orthonormal and $I_{3, 1}$ is diagonal, $U I_{3, 1} U^\top$
is also an eigendecomposition of $\Pi$ where the fixed
points of $\Pi$ are mapped to the eigenvectors $e_1$ and $e_4$
while the cycles of order two are mapped to the eigenvectors  
$\tfrac{1}{\sqrt{2}}(e_{2} + e_3)$ and $\tfrac{1}{\sqrt{2}}(e_{2} -
e_3)$; here $e_i$ denote the $i^\mathrm{th}$ basis vector in $\mathbb{R}^{4}$.
\end{proof}

\begin{theorem}[Generalization to $K > 2$]
\label{theorem2}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell)}\})$ be an instance of a
PABM with $K \geq 1$ blocks and latent vectors $\{\lambda^{(k \ell)}
\colon 1 \leq k \leq K, 1 \leq \ell \leq K\}$. Then there exists a block diagonal matrix
$X \in \mathbb{R}^{n \times K^2}$ defined by the $\{\lambda^{(k \ell)}\}$ and a $K^2 \times
K^2$ {\em fixed} orthonormal matrix $U$ such 
that $A \sim \mathrm{GRDPG}_{K (K+1) / 2, K (K-1) / 2}(XU)$.
\end{theorem}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Community membership identification (up to permutation).
\item
  Parameter estimation (estimating \(\lambda^{(k \ell)}\)'s).
\end{enumerate}

@@

\citet*{noroozi2019estimation} used Sparse Subspace Clustering (SSC) 
\cite{5206547} for community detection in the PABM. 
The SSC algorithm can be described as follows: 
Given \(X \in \mathbb{R}^{n \times d}\) with vectors
\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
\(c_i = \arg\min_{c} ||c||_1\) subject to \(x_i = X c\) and
\(c^{(i)} = 0\) is solved for each \(i = 1, ..., n\). The solutions are
collected collected into matrix
\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
construct an affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
perfectly on one of \(K\) subspaces, \(B\) describes an undirected graph
consisting of \(K\) disjoint subgraphs, i.e., \(B_{ij} = 0\) if
\(x_i, x_j\) lie on different subspaces. If \(X\) instead represents
points near \(K\) subspaces with some noise, a final graph partitioning
step may be required (e.g., edge thresholding or spectral clustering).

@@

For any edge probability matrix \(P\) for the PABM such that the rows
and columns are organized by community, the \(kl\)\textsuperscript{th}
block is an outer product of two vectors, i.e.,
\(P^{(k \ell)} = \lambda^{(k \ell)} (\lambda^{(\ell k)})^\top\). Therefore, given
\(P^{(k \ell)}\), \(\lambda^{(k \ell)}\) and \(\lambda^{(\ell k)}\) are solvable
up to multiplicative constant using singular value
decomposition. More specifically, 
let \(P^{(k \ell)} = (\sigma^{(k \ell)})^2 u^{(k \ell)} (v^{(k \ell)})^\top\)
be the singular value decomposition of \(P^{(k \ell)}\).
\(u^{(k \ell)} \in \mathbb{R}^{n_k}\) and 
\(v^{(k \ell)} \in \mathbb{R}^{n_\ell}\) are vectors
and \(\sigma^{(k \ell)}\) is a scalar. 
Then \(\lambda^{(k \ell)} = s_1 u^{(k \ell)}\)
and \(\lambda^{(\ell k)} = s_2 v^{(k \ell)}\) 
for unidentifiable $s_1 s_2 = (\sigma^{(k \ell)})^2$.
Given the adjacency matrix \(A\)
instead of edge probability matrix \(P\), we can simply use plug-in
estimators by taking the SVD of each $A^{(k \ell)}$. 
Since each $\lambda^{(k \ell)}$ is not strictly identifiable,
we instead estimate each $\tilde{\lambda}^{(k \ell)} = \sigma^{(k \ell)} u^{(k \ell)}$.

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{PABM parameter estimation.}
  \KwData{Adjacency matrix $A$, community assignments $1, ..., K$}
  \KwResult{PABM parameter estimates $\{\hat{\lambda}^{(k \ell)}\}_K$.}
  Arrange the rows and columns of $A$ by community such that each 
  $A^{(k \ell)}$ block consists of estimated edge probabilities between 
  communities $k$ and $l$.\;
  \For {$k, l = 1, ..., K$, $k \leq l$} {
    Compute $A^{(k \ell)} = U \Sigma V^\top$, the SVD of the $kl^{th}$ 
    block.\;
    Assign $u^{(k \ell)}$ and $v^{(k \ell)}$ as the first columns of $U$ and $V$. 
    Assign $(\sigma^{(k \ell)})^2 \leftarrow \Sigma_{11}$.\;
    Assign $\hat{\lambda}^{(k \ell)} \leftarrow \pm \sigma^{(k \ell)} u^{(k \ell)}$ and 
    $\hat{\lambda}^{(\ell k)} \leftarrow \pm \sigma^{(k \ell)} v^{(k \ell)}$.
  }
\end{algorithm}

\begin{theorem}
\label{theorem6}
Let each $\tilde{\lambda}^{(k \ell)}$ be the popularity vector derived from its corresponding $P^{(k \ell)}$, and let $\hat{\lambda}^{(k \ell)}$ be its estimate from $A^{(k \ell)}$ (Alg. 3). Then if $n \rho_n = \omega \big( (\log n)^{4 c} \big)$,

\begin{equation} \label{eq:thm6}
\max_{k, l \in \{1, ..., K\}} 
\|\hat{\lambda}^{(k \ell)} - \tilde{\lambda}^{(k \ell)}\|_{\infty} = 
O\bigg(\frac{(\log n_k)^c}{\sqrt{n_k}} \bigg)
\end{equation}

@@

For each simulation, community labels are drawn from a multinomial
distribution, the popularity vectors \(\{\lambda^{(k \ell)}\}_K\) are drawn
from two types of joint distributions depending on whether \(k = l\),
the edge probability matrix \(P\) is constructed using the popularity
vectors, and finally an unweighted and undirected adjacency matrix \(A\)
is drawn from \(P\). OSC is then used for community detection, and this
method is compared against SSC \cite{noroozi2019estimation,soltanolkotabi2014}
and MM \cite{igraph, 307cbeb9b1be48299388437423d94bf1}. 
True community labels are used
with Algorithm 3 to estimate the popularity vectors
\(\{\lambda^{(k \ell)}\}_K\), and this method is then compared against an
MLE-based estimator described by \citet{noroozi2019estimation} and
\citet{307cbeb9b1be48299388437423d94bf1}.

@@

We also define two types of parameter estimation error. 
First, we estimate the popularity vectors directly and compute the RMSE. 
In this case, the "true" popularity vectors are derived from 
taking the SVD of each edge probability block $P^{(k \ell)}$ 
to avoid the unidentifiable multiplicative constants.

\[\text{RMSE}(\{\hat{\lambda}^{(k \ell)}\}_K, \{\lambda^{(k \ell)}\}_K) = 
\sqrt{\frac{1}{n} \sum_{k < l} \min_{s = \pm 1} 
\|s \hat{\lambda}^{(k \ell)} - \lambda^{(k \ell)} \|_2^2}\]

We can also avoid the unidentifiable multiplicative constant more directly 
by reconstructing each 
$\hat{P}^{(k \ell)} = \hat{\lambda}^{(k \ell)} (\hat{\lambda}^{(\ell k)})^\top$, 
which we use to define another parameter estimation error.

\[\text{RMSE}(\hat{P}, P) =
\sum_{k,l} \sqrt{\frac{1}{n_k n_\ell} \|P^{(k \ell)} - \hat{P}^{(k \ell)}\|_F^2}\]

@@

In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
multinomial distribution with mixture parameters
\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(k \ell)}\}_K\) according
to the drawn community labels, \(P\) was constructed using the drawn
\(\{\lambda^{(k \ell)}\}_K\), and \(A\) was drawn from \(P\) by
\(A_{ij} \indep \Bernoulli(P_{ij})\). Each simulation has
a unique edge probability matrix \(P\).

@@

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = 1 / K\) for \(k = 1, ..., K\), (i.e.,
  each community label has an equal probability of being drawn)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
  \(k \neq l\)
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = \frac{k^{-1}}{\sum_{l=1}^K l^{-1}}\)
  for \(k = 1, ..., K\)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(k \ell)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
  \(k \neq l\)
\end{itemize}

@@

\hypertarget{proofs}{%
\section{Proofs of Theorem~\ref{theorem4} and Theorem~\ref{theorem5}}}
Let \(V_n\) and \(\hat{V}_n\)
be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
$K^2$ largest eigenvalues (in modulus), respectively. 
We first state an important technical lemma for bounding the maximum
$\ell_2$ norm difference between the rows of $\hat{V}_n$ and
$V_n$. See \cite{cape_biometrika} and 
\cite[Lemma~5]{rubindelanchy2017statistical} for a proof. 
\begin{lemma}
  \label{lem:technical}
Let $A \sim \mathrm{PABM}(\{\lambda^{(k \ell \ell)}\}_{K})$ be a $K$-blocks
PABM graph on $n$ vertices and let \(V_n\) and \(\hat{V}_n\)
be the $n \times K^2$ matrices whose columns are the eigenvectors of \(P\) and \(A\) corresponding to the
$K^2$ largest eigenvalues in modulus, respectively.
Let \((v_n^{(i)})^\top\) and \((\hat{v}_n^{(i)})^\top\) denote the $i$th 
row of \(V_n\) and \(\hat{V}_n\), respectively. 
Then there exists a constant $c > 1$ and an orthogonal matrix $W$ such
that, with high probability,
$$\max_{i} \|W \hat{v}_n^{(i)} - v_n^{(i)}\|  = O_P \Big(\frac{\log^{c}n}{n \sqrt{\rho_n}} \Big).$$
In particular we can take $c = 1 + \epsilon$ for any $\epsilon > 0$. 
\end{lemma}

@@

which is a symmetric \((n_k + n_\ell) \times (n_k + n_\ell)\) matrix. The
eigendecomposition of \(M^{(k \ell)}\) is then

\[M^{(k \ell)} = 
\begin{bmatrix} u^{(k \ell)} & -u^{(k \ell)} \\ v^{(k \ell)} & v^{(k \ell)} \end{bmatrix} \times 
\begin{bmatrix} \sigma^2_{kl} & 0 \\ 0 & -\sigma^2_{kl} \end{bmatrix} \times
\begin{bmatrix} u^{(k \ell)} & -u^{(k \ell)} \\ v^{(k \ell)} & v^{(k \ell)} \end{bmatrix}^\top\]
Thus treating \(M^{(k \ell)}\) as the edge probability matrix of a GRDPG, we
have latent positions in \(\mathbb{R}^2\) given by the $(n_k + n_{\ell}) \times 2$ matrix
\[\Lambda^{(k \ell)} = \begin{bmatrix} 
  \sigma_{k \ell} u^{(k \ell)} & \sigma_{k \ell} u^{(k \ell)} \\ 
  \sigma_{k \ell} v^{(k \ell)} & -\sigma_{k \ell} v^{(k \ell)} 
\end{bmatrix} = 
\begin{bmatrix} 
  \lambda^{(k \ell)} & \lambda^{(k \ell)} \\ 
  \lambda^{( \ell k)} & -\lambda^{( \ell k)} 
\end{bmatrix}.\]
Now consider
\[\hat{M}^{(k \ell)} = \begin{bmatrix} 0 & A^{(k \ell)} \\ A^{(\ell k)} & 0 \end{bmatrix}\]
% Then \(\hat{M}^{(kl)} = M^{(kl)} + E'\) where
